---
title: "papaja"
author: "Matt"
date: "1/31/2019"
output: 
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)
```

`papaja` is an R package for **P**reparing **APA** **J**ournal **A**rticles. It allows you to write in Rmarkdown (.rmd file), and then compile your document as a .pdf in APA format (double-spaced, all the APA style things)

- full documentation for using papaja [https://crsh.github.io/papaja_man/](https://crsh.github.io/papaja_man/)
- github repository for papaja [https://github.com/crsh/papaja](https://github.com/crsh/papaja)

**Important note: papaja will not work if you attempt to use it inside the R project for your website, you need to create a new R project folder**

## Installation

There is one small complication with using `papaja`. You need to have TEX (or Latex) installed on your computer before it will work. 

Latex is a powerful typesetting program that is often used to produce .pdfs. R Markdown can output .rmd files to .pdf when you have latex installed on your computer.

### Installing Latex

- Mac users should install MacTex [https://tug.org/mactex/](https://tug.org/mactex/)
  - note there are two installations called Mactex download (the full installation, it's big), and Basic Tex (smaller download). I know that the Mactex (big download) works. You can try the smaller download, but I don't know if it works.
  
- Windows users should install MikTex [https://miktex.org](https://miktex.org)
  - I have never tried to do this on windows, I'm assuming it works.
  
- R-studio cloud. A latex distribution is already installed if you use R-studio cloud. However, there is one step you need to take in order to get papaja to work, and that is to include the `apa.cls` file in your main project directory.
  - Below is a link to the `apa.cls` file. A reminder that you will need to create a new R project on R-studio cloud. Then, copy this file into that new R project. 
  - [https://github.com/CrumpLab/psyc7709/blob/master/Presentations/apa6.cls](https://github.com/CrumpLab/psyc7709/blob/master/Presentations/apa6.cls)

### Installing papaja

`papaja` is not available on CRAN, so you have to download it from github.

- The following code uses the `devtools` library to install from github. If you don't have `devtools` installed, first install it using R-studio (it is on CRAN, so you can install it from the packages tab)
- Then run the following line in the console of R-studio.

```{r, eval=FALSE}
devtools::install_github("crsh/papaja")
```

## Using papaja

We will go over papaja in class. Follow these steps to check if papaja works on your computer.

1. Create a new R project. **Do not attempt the next steps inside the project file for your website, it will not work, you need to start fresh with a new R project folder**.
2. Select the green plus (top-left corner), and choose `R Markdown...` to create a new .rmd file.
3. This will open a window, look at the column on the left and choose from template, find the "APA article (6th edition)" option, and choose that.
4. This will open up a default `papaja` template .rmd file. 
5. Knit the file.
6. If everything works, then you will create a new .pdf with showing a bare minimum APA-style manuscript.


