---
title: "Example Analysis"
author: "Matt"
date: "1/31/2019"
output: 
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)
```

This example is taken from [https://crumplab.github.io/statisticsLab/lab-10-factorial-anova.html#r-10](https://crumplab.github.io/statisticsLab/lab-10-factorial-anova.html#r-10). The data file is `stroop_stand.csv`, and can be found here [https://github.com/CrumpLab/statisticsLab/tree/master/data](https://github.com/CrumpLab/statisticsLab/tree/master/data)

## Load the data and libraries you will use

```{r}
library(data.table)
library(dplyr)
library(ggplot2)
all_data <- fread("data/stroop_stand.csv")
```

## Pre-processing

Pre-processing can include many steps. Here, we convert the data from wide to long format, and make a new data frame.

```{r}
RTs <- c(as.numeric(unlist(all_data[,1])),
         as.numeric(unlist(all_data[,2])),
         as.numeric(unlist(all_data[,3])),
         as.numeric(unlist(all_data[,4]))
         )

Congruency <- rep(rep(c("Congruent","Incongruent"),each=50),2)
Posture <- rep(c("Stand","Sit"),each=100)
Subject <- rep(1:50,4)

stroop_df <- data.frame(Subject,Congruency,Posture,RTs)
```

### Checks

It is important to check the data you are analyzing before you analyze. The checks you make depend on the questions you are trying to answer. Here we check the number of subjects in each condition.

This is a 2x2 repeated measures design, each subject should have 4 means

```{r}
num_subjects <- stroop_df %>%
                  group_by(Subject) %>%
                  summarise(counts = length(Subject))
total_subjects <- length(num_subjects$Subject)
total_has_four <- num_subjects$counts == 4
sum(total_has_four) == total_subjects

```

Are there any subjects with huge mean RTs? Could indicate something weird happened.

```{r}
hist(stroop_df$RTs)
```

### Exclusion

If you are going to exlcude subjects from analysis, then justify your exclusion criterion, and then eliminate subject data from the data frame. We will not exlcude any subjects here.

```{r}
# no exclusions
```

## Analysis

Now you can begin analysis. This experiment asked whether the Stroop effect (difference between mean incongruent and mean congruent), depends on whether people are sitting or standing. Let's get the means, make a tbale, then plot the data.

### table of 2x2 means

```{r}
overall_means <- stroop_df %>%
                  group_by(Posture,Congruency) %>%
                  summarise(meanRT = mean(RTs),
                            SEMRT = (sd(RTs)/sqrt(length(RTs))))

# make a table of overall means
knitr::kable(overall_means)
```

### plot of 2x2 means

```{r}
ggplot(overall_means, aes(x=Posture,
                          y=meanRT, 
                          group=Congruency,
                          fill=Congruency))+
  geom_bar(stat="identity",position="dodge")+
  theme_classic(base_size=12)+
  ylab("Mean Reaction Time (ms)")+
  geom_errorbar(aes(ymin=meanRT-SEMRT,
                    ymax=meanRT+SEMRT),
                position=position_dodge(width=0.9),
                width=.2,
                color="black")+
  coord_cartesian(ylim=c(750,1000))
```

### Repeated measure ANOVA

```{r}
# Make sure Subjecdt is  a factor
stroop_df$Subject <-  as.factor(stroop_df$Subject)

aov_out <- aov(RTs~Posture*Congruency + Error(Subject/(Posture*Congruency)), stroop_df)
#print summary of ANOVA table
summary(aov_out)


# prints a nicer ANOVA table
summary_out <- summary(aov_out)
library(xtable)
knitr::kable(xtable(summary_out))

# prints means for each effect
print(model.tables(aov_out,"means"), format="markdown")

```
 
### Example Write-Up
 
 We submitted the mean reaction times for each subject in each condition to a 2 (Congruency: congruecnt vs. incongruent) x 2 (Posture: Standing vs. Sitting) repeated measures ANOVA.

There was a main effect of Congruency, F (1, 49) = 342.45, MSE = 1684.39, p < 0.001. Mean reaction times were slower for incongruent (922 ms) than congruent groups (815 ms).

There main effect of Posture was significant, F (1, 49) = 7.33, MSE = 4407.09, p =.009. Mean reaction times were slower for sitting (881 ms) than standing groups (855 ms).

The two-way interaction between Congruency and Posture was significant, F (1, 49) = 8.96, MSE = 731.82, p < 0.004. The Stroop effect was 23 ms smaller in the standing than sitting conditions.





