---
title: "Statistics"
author: "Matt"
date: "1/31/2019"
output: 
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)
```

**The most important thing about choosing statistics, is being able to justify the statistics that you choose.**

There are many tools in the toolbox. There are many recommendations about what to do or not to do. It is a good idea to understand the statistics that you use, so that you can justify why you are using them for your analysis.

## t-tests

### one sample t-test

Set `mu` to test against a population value

```{r}
x <- c(1,4,3,2,3,4,3,2,3,2,3,4)

#non-directional
t.test(x, mu = 2)
t.test(x, mu = 2, alternative = "two.sided") # the default

# directional
t.test(x, mu = 2, alternative = "greater")
t.test(x, mu = 2, alternative = "less")
```

### paired-sample t-test

Set `paired=TRUE`

```{r}
x <- c(1,4,3,2,3,4,3,2,3,2,3,4)
y <- c(3,2,5,4,3,2,3,2,1,2,2,3)

#non-directional
t.test(x, y, paired=TRUE)
t.test(x, y, paired=TRUE, alternative = "two.sided") # the default

# directional
t.test(x, y, paired=TRUE, alternative = "greater")
t.test(x, y, paired=TRUE, alternative = "less")
```

### Independent sample t-test

Note: Default is Welch test, set `var.equal=TRUE` for t-test

```{r}
x <- c(1,4,3,2,3,4,3,2,3,2,3,4)
y <- c(3,2,5,4,3,2,3,2,1,2,2,3)

#non-directional
t.test(x, y, var.equal=TRUE)
t.test(x, y, var.equal=TRUE, alternative = "two.sided") # the default

# directional
t.test(x, y, var.equal=TRUE, alternative = "greater")
t.test(x, y, var.equal=TRUE, alternative = "less")
```

### printing 

```{r}
library(broom)

t_results <- t.test(x, y, var.equal=TRUE)
knitr::kable(tidy(t_results))

```

### writing

The contents of R variables can be written into R Markdown documents. 

t(`r round(t_results$parameter, digits=2)`) =`r round(t_results$statistic, digits=2)`, p = `r round(t_results$p.value, digits=3)`

```{r}
# write your own function
report_t <- function(x){
  return(paste(c("t(",round(x$parameter, digits=2),") ",
        "= ",round(x$statistic, digits=2),", ",
        "p = ",round(x$p.value, digits=3)), collapse=""))
}
report_t(t_results)
t_results
```


The results of my t-test were `r report_t(t_results)`.


## ANOVA

Requires data to be in long-format.

```{r}
# example creation of 2x2 data 
Subjects <- rep(1:10,each=4)
Factor1 <- rep(rep(c("A","B"), each = 2), 10)
Factor2 <- rep(rep(c("1","2"), 2), 10)
DV <- rnorm(40,0,1)

all_data <- data.frame(Subjects = as.factor(Subjects),
                       DV,
                       Factor1,
                       Factor2)
```

### between-subjects 1 factor

```{r}

# run anova
aov_out <- aov(DV~Factor1, all_data)

# summary
summary(aov_out)

# kable xtable printing
library(xtable)
knitr::kable(xtable(summary(aov_out)))

# print means
print(model.tables(aov_out,"means"), format="markdown")
```

### between-subjects 2 factor

```{r}

# run anova
aov_out <- aov(DV~Factor1*Factor2, all_data)

# summary
summary(aov_out)

# kable xtable printing
library(xtable)
knitr::kable(xtable(summary(aov_out)))

# print means
print(model.tables(aov_out,"means"), format="markdown")
```

### repeated-measures 1 factor

```{r}

# run anova
aov_out <- aov(DV~Factor1 + Error(Subjects/Factor1), all_data)

# summary
summary(aov_out)

# kable xtable printing
library(xtable)
knitr::kable(xtable(summary(aov_out)))

# print means
print(model.tables(aov_out,"means"), format="markdown")
```


### repeated-measures 2 factor

```{r}

# run anova
aov_out <- aov(DV~Factor1*Factor2 + Error(Subjects/(Factor1*Factor2)), all_data)

# summary
summary(aov_out)

# kable xtable printing
library(xtable)
knitr::kable(xtable(summary(aov_out)))

# print means
print(model.tables(aov_out,"means"), format="markdown")
```

### papaja

```{r}
library(papaja)

apa_stuff <- apa_print.aov(aov_out)
```

The main effect for factor 1 was, `r apa_stuff$statistic$Factor1`. The main effect for factor 2 was, `r apa_stuff$statistic$Factor2`. The interaction was, `r apa_stuff$statistic$Factor1_Factor2`

## Linear Regression

```{r}
lm(DV~Factor1, all_data)
summary(lm(DV~Factor1, all_data))
```

```{r}
lm(DV~Factor1+Factor2, all_data)
summary(lm(DV~Factor1+Factor2, all_data))
```

```{r}
lm(DV~Factor1*Factor2, all_data)
summary(lm(DV~Factor1*Factor2, all_data))
```





## Randomization Test

```{r}
A <- c(1,2,3,4,5,6,7,8,9,10)
B <- c(2,4,6,8,10,12,14,16,18,20)

all <- c(A,B)

mean_difference <- c()
for(i in 1:10000){
  shuffle <- sample(all)
  newA <- shuffle[1:10]
  newB <- shuffle[11:20]
  mean_difference[i] <- mean(newB)-mean(newA)
}

observed <- mean(B)-mean(A)
length(mean_difference[mean_difference >= observed])/10000

library(EnvStats)

twoSamplePermutationTestLocation(x=B, y=A, fcn = "mean", 
                                 alternative = "greater", 
                                 mu1.minus.mu2 = 0, 
                                 paired = FALSE, 
                                 exact = FALSE, 
                                 n.permutations = 10000, 
                                 seed = NULL, 
                                 tol = sqrt(.Machine$double.eps))

```

## Correlation

```{r}
x <- rnorm(10,0,1)
y <- rnorm(10,0,1)
cor(x,y)

ranks1 <- sample(1:10) 
ranks2 <- sample(1:10)
cor(ranks1,ranks2, method = "spearman")
```

## Chi-square

```{r}
M <- as.table(rbind(c(762, 327, 468), c(484, 239, 477)))
dimnames(M) <- list(gender = c("F", "M"),
                    party = c("Democrat","Independent", "Republican"))
(Xsq <- chisq.test(M))
```

## binomial test

```{r}
number_of_successes <- 100
trials <- 150
binom.test(x=number_of_successes, n=trials, p = 0.5,
           alternative = "two.sided",
           conf.level = 0.95)
```

## Post-hoc tests

Many kinds of post-hoc tests can be done in R using base R functions, or functions from other packages. 

At the same time, it is common for post-hoc comparison functions to be limited and specific in their functionality. So, if you can't find an existing function to complete the analysis you want, you may have to modify, extend, or write your own function to do the job. 

### pairwise t-tests

Performs all possible t-tests between the levels of a grouping factor. Can implement different corrections for multiple comparisons.

```{r}
df <- airquality #loads the airquality data from base R
df$Month <- as.factor(df$Month)

# default uses holm correction
pairwise.t.test(df$Ozone,df$Month)

# use bonferroni correction
pairwise.t.test(df$Ozone,df$Month, p.adj="bonf")
```

### tukey HSD test

notes: the warpbreaks data.frame should be available in base R, you do not need to load it to be able to use it.

Doesn't work with repeated measures ANOVAs.

```{r}
summary(fm1 <- aov(breaks ~ wool + tension, data = warpbreaks))

TukeyHSD(fm1, "tension", ordered = TRUE)
```

### Fishers LSD

Explanation of Fisher's least significant difference test.
[https://www.utd.edu/~herve/abdi-LSD2010-pretty.pdf](https://www.utd.edu/~herve/abdi-LSD2010-pretty.pdf)

```{r}
library(agricolae)

data(sweetpotato)
model <- aov(yield~virus, data=sweetpotato)
out   <- LSD.test(model,"virus", p.adj="bonferroni")

```

## Linear Contrasts

```{r}
# Create data for a one-way BW subjects, 4 groups
A <- rnorm(n=10, mean=100, sd=25)
B <- rnorm(n=10, mean=120, sd=25)
C <- rnorm(n=10, mean=140, sd=25)
D <- rnorm(n=10, mean=80, sd=25)

DV <- c(A,B,C,D)
Conditions <- as.factor(rep(c("A","B","C","D"), each=10))
df <- data.frame(DV,Conditions)

# one-way ANOVA
aov_out <- aov(DV~Conditions, df)
summary(aov_out)

# look at the order of levels
levels(df$Conditions)

# set up linear contrasts
c1 <- c(.5, -.5, -.5, .5) # AD vs. BC
c2 <- c(0, 0, 1, -1) # C vs. D
c3 <- c(0, -1, 1, 0) # B vs. C

# create a contrast matrix
mat <- cbind(c1,c2,c3)

# assign the contrasts to the group
contrasts(df$Conditions) <- mat

# run the ANOVA
aov_out <- aov(DV~Conditions, df)

# print the contrasts, add names for the contrasts
summary.aov(aov_out, split=list(Conditions=list("AD vs. BC"=1, "C vs. D" = 2, "B vs. C"=3))) 
```

