# The sf package

**Author: Kamal Abdelrahman

```{r, include=FALSE}
library(scales)
library(ggplot2)
library(sf)
library(data.table)
```



# Goals With Spatial Features 
1. Getting to Know GIS

2. Introducing the sf package  

3. Example

# Getting To Know GIS
Geographic Information Systems - Framework of capturing, storing, manipulating, analyzing, managing, and presenting spatial data
  + Features - The geometry of a shape, bits of points that comprise its structure on given place on Earth
  
  + Spatial Data - Data identifies geoprahical location
  
  + Attribute Data - Data that identifies characteristics of locations
  
  + Shapefiles - Geometric shape of spatial data
      + **NOTE** - MUST BE KEPT IN SAME FOLDER of supporting files
  
  + Choropleth - Thematic map with areas of study site shaded in proportion to values of an area

# Introducing the sf package

The sf, short for **simple features**, package allows analysts to process and analyze geospatial data in R, noted by tibbles/data.frames with the **geometry column**

  * Importing/Exporting shapefiles
  
  * Creating choropleths
  
  * Works well with ggplot2 syntax 

# Example 

You're analyzing disparities of income levels across New York City and want to use a choropleth to visualize that

## What will we need?

  * New York City Shapefile (Community District Level)
  * New York City Median Incomes (Community District Level)

## How will we actually make one?
  1. Read in our shapefile, and dataset
  2. Merge them together 
  3. Visualize the map using ggplot syntax
  
## Step 1 - Reading In Shapefile/Data

Here is where is where read in the shapefile and dataset

```{r eval=FALSE, include=TRUE}
install.packages("sf")
```
```{r eval=FALSE, include=TRUE}
library(sf)
#Reading From File
NycCommunityDistricts <- read_sf("/filpath/shapefilename.shp")
#Reading From Website
NycCommunityDistricts <- read_sf("https://data.cityofnewyork.us/api/geospatial/yfnk-k7r4?method=export&format=GeoJSON")
#Load in Income Levels
NYCIncome <- fread("https://raw.githubusercontent.com/kamalabdel97/FinalPresentation/master/Median%20Incomes.csv")
```

```{r eval=TRUE, include=FALSE}
#Reading From File
library(sf)
NycCommunityDistricts <- read_sf("https://data.cityofnewyork.us/api/geospatial/yfnk-k7r4?method=export&format=GeoJSON")
NYCIncome <- fread("https://raw.githubusercontent.com/kamalabdel97/FinalPresentation/master/Median%20Incomes.csv")
```

```{r eval=TRUE, include=FALSE}
NYCIncome <- subset(NYCIncome, nchar(NYCIncome$Fips) == 3 & NYCIncome$`Household Type` == "All Households" & NYCIncome$TimeFrame == "2017")
```

`read_sf` - Reads in simple features

# New York Stand Up!

```{r echo=FALSE, fig.align='center'}
ggplot() + geom_sf(data = NycCommunityDistricts)
```


# Step 2 - Joining the data to the shapefile

```{r eval=TRUE, include=TRUE}
NycCommunityDistrictsIncome <- merge(NycCommunityDistricts, NYCIncome, by.x = "boro_cd", by.y = "Fips", all.x = T)
```

`NycCommunityDistricts` - The shapefile of the New York City community districts

`NYCIncome` - The attribute data contain the income levels

# Step 3 - Making the choropleth

```{r eval=FALSE, include=TRUE}
ggplot() + 
  geom_sf(data = NycCommunityDistrictsIncome,
  aes(fill = Income)) 
```

`ggplot` - Invokes the ggplot syntax for data visualization

`geom_sf` - Visualizes the shapefile feature
* `NycCommunityDistrictsIncome` - The shapefile dataframe containing BOTH the shapefile (spatial) and income levels (attribute) 

* `Income` - The statistical value in the shapefile dataframe for each community district

# Final Map!
```{r echo=FALSE, fig.align='center'}
ggplot() + 
  geom_sf(data = NycCommunityDistrictsIncome, aes(fill = Income), color = NA)+
  scale_fill_gradient(low = rgb(204, 209, 205, maxColorValue=255), high = rgb(0, 100, 46, maxColorValue=255))
```