---
title: "Making Packages in R"
subtitle: "Using pkgdown and roxygen2"
author: "Matthew Crump"
date: "2019/4/29 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

class: inverse, center, middle

# How to use pkgdown and roxygen2 to build and share R packages

---

# What is pkgdown and roxygen2?

## pkgdown

- a library to convert the documentation for your R package into an an awesome website

## roxygen2

- a library to make writing documentation for your R package faster and easier, and more betterly.

---

# pkgdown

## How to install

```{r, eval = FALSE}
# Install release version from CRAN
install.packages("pkgdown")
```

## More info about pkgdown

- (https://pkgdown.r-lib.org)[https://pkgdown.r-lib.org]
- (A vignette on using pkgdown)[https://pkgdown.r-lib.org/articles/pkgdown.html]

---

# roxygen2

## How to install

```{r, eval = FALSE}
# Install release version from CRAN
install.packages("roxygen2")
```

## helpful tutorials on getting started with roxygen2

- (http://r-pkgs.had.co.nz/man.html)[http://r-pkgs.had.co.nz/man.html]

---

class: inverse, center, middle

# Show me an example of why I should bother with this

---

# Ok, but first

## R-packages

Hadley Wickham wrote a wonderful and free book on making R packages, and it is worth looking at:

(http://r-pkgs.had.co.nz)[http://r-pkgs.had.co.nz]

---

# pkgdown example

I recently made an R package for calculating statistical power (for particular designs).

I compiled the packaged using pkgdown, let's a take a look:

[https://crumplab.github.io/conflictPower/](https://crumplab.github.io/conflictPower/)

---

# compiling a package with pkgdown

1. Make an R package (e.g., start with the R template for packages)
2. Install pkgdown
3. Run the following line when you want to compile your package using pkgdown:

```{r, eval=FALSE}
pkgdown::build_site()
```

4. That's it...pkgdown creates a new docs folder in your R project folder that contains a website for your R package. If you upload the folder to github, and use github pages to serve the website through the docs folder, then you can share it on the web.

---

# Writing documentation with roxygen2

"nothing here yet..."

---
