
# Network Analysis tutorial

**Author: Jiaxin Deng**

## Brief introduction to network analysis

A network is a set of nodes connected by a set of edges.

Several packages are used in the network analysis, including `network`, `statnet`, `igraph` and `qgraph`.

`qgraph` was developed in the context of psychometrics approach by Dr. Sacha Epskamp and colleagues in 2012. For more details, please click this following link for the paper published in *Journal of Statistical Softare*:

<https://www.jstatsoft.org/article/view/v048i04>

This package can create graphs to visualize the statistics in different layout modes based on different correlation matrices, such as polychoric correlation, partial correlation.

## Example code

Here is the following steps to conduct a network analysis using `qgraph`.

Take `big5` data as an example. This is a dataset of the Big five personality traits assessed on 500 psychology students.

Firstly, `qgraph` package should be activated using `library()`

```{r library()}
library(qgraph)
```

And then, data need to be imported in the current R project.

```{r get big5 data}
data(big5)
```


To creat the graph is basically to use `qgraph()`, such as:

```{r cor big5}
qgraph<-qgraph(cor(big5))
```


But it should be noted that the input in the `qgraph()` can be a weight matrix or an edgelist.

Thus, if you want to creat the association network, `cor()`/`cor_auto()` can be used to creat the matrix first.

Also, you can use `groups` to indicate which nodes belong together, such as:

```{r groups big5}
data("big5groups")
qgraph(cor(big5), groups=big5groups)
```

Besides, you can use some additional arguments to customize your representing graph.

you can use `layout` to change the representation, such as:

```{r layout big5}
qgraph(cor(big5), groups=big5groups,layout= "spring")
```

```{r layout2 big5}
qgraph(cor(big5), groups=big5groups,layout= "circle")
```





## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
